---
title: "One Gridpoint"
author: "Patricia Gribi"
date: "2024-03-19"
output: html_document
---

# Cumulative Water Deficits over Time for one Gridpoint

## Data Wrangling

### Overview

The data wrangling started with reading the data and loading the required libraries. The CMIP6ng data structure is organised in such a way that each variable is contained in a separate file. Therefore, each variable had to be read in individually. The data format is netcdf and contains a global grid. However, this is kept in a purely positive format (often preferred for climate data). A grid point was extracted for every variable. Also some unit conversions had to be performed. Variables such as precipitation and evapotranspiration were converted to the desired mm day-1 format (equivalent to kg m-2). The temporal resolution of the individual variables is daily or monthly. Linear interpolation was therefore applied to the variables that are only available monthly in order to obtain the same resolution for all variables. Then, evapotranspiration and potential evapotranspiration were calculated over time for a specific grid point. Furthermore, snow was modelled. Thanks to all these steps, the cwd algorithm could finally be applied to the data.

### Library Loading

```{r, echo=FALSE, message=FALSE}

# load libraries
source("../R/load_libraries.R")
load_libraries()

```

### Data Reading

```{r}

# reading data from repo
evspsbl <- read.csv(paste0(here::here(),"/data/gridpoint_evspsbl.csv"), header=TRUE)
pr <- read.csv(paste0(here::here(),"/data/gridpoint_pr.csv"), header=TRUE)
tas <- read.csv(paste0(here::here(),"/data/gridpoint_tas.csv"), header=TRUE)
rlus <- read.csv(paste0(here::here(),"/data/gridpoint_rlus.csv"), header=TRUE)
rlds <- read.csv(paste0(here::here(),"/data/gridpoint_rlds.csv"), header=TRUE)
rsds <- read.csv(paste0(here::here(),"/data/gridpoint_rsds.csv"), header=TRUE)
rsus <- read.csv(paste0(here::here(),"/data/gridpoint_rsus.csv"), header=TRUE)

```

### Column Renaming and Unit Conversions

This research project aims to consistently represent the variables across all data sets with the following units:

-   **Date**: year-month-day (xxxx-xx-xx)
-   **Precipitation** and **(potential) Evapotranspiration**: in mm day-1 (equivalent to kg m-2)
-   **Temperature (tas)**: in degree Celsius (°C)
-   **Atmospheric Pressure**: in Pascal (Pa)
-   **Radiation**: in W m-2

Therefore some unit conversions had to be conducted.

```{r}

# evapotranspiration
colnames(evspsbl) <- c("date", "evapotranspiration") # column renaming
evspsbl$evapotranspiration <- evspsbl$evapotranspiration * 86400 # conversion to mm day-1

# precipitation
colnames(pr) <- c("date", "precipitation") # column renaming
pr$precipitation <- pr$precipitation * 86400 # conversion to mm day-1

# temperature
colnames(tas) <- c("date", "temperature") # column renaming
tas$temperature <- tas$temperature - 273.15 # conversion to °C

# radiation
colnames(rlus) <- c("date", "up_longwave_radiation") # column renaming
colnames(rlds) <- c("date", "down_longwave_radiation") # column renaming
colnames(rsds) <- c("date", "down_shortwave_radiation") # column renaming
colnames(rsus) <- c("date", "up_shortwave_radiation") # column renaming

# merge radiation variables into one dataframe
radiation_df <- merge(merge(merge(rlus, rlds, by = "date"), rsds, by = "date"), rsus, by = "date")

# calculate net radiation and add to dataframe
radiation_df$net_radiation <- (radiation_df$down_shortwave_radiation - radiation_df$up_shortwave_radiation) + (radiation_df$down_longwave_radiation- radiation_df$up_longwave_radiation)

net_radiation_df <- select(radiation_df, date, net_radiation)
write.csv(net_radiation_df, "../data/net_radiation_df.csv", row.names = FALSE)


```

In order to calculate the potential evapotranspiration the net radiation was needed. The net radiation is calculated as net shortwave radiation + net longwave radiation (Brönnimann 2018, p. 89).

### Time-Range Adjustments

To obtain a daily resolution for all variables, linear interpolation was applied to monthly available variables.

```{r}

# creation of new dataframe to calculate pet

source("../R/exclude_time_range.R")

# precipitation subset selection
adj_pr <- exclude_time_range(pr, as.Date("2014-11-22"), as.Date("2014-11-30") )

# temperature subset selection
adj_tas <- exclude_time_range(tas, as.Date("2014-11-22"), as.Date("2014-11-30") )

# creation of the date sequence
date_sequence <- seq(as.Date("2014-12-01"), as.Date("2100-10-31"), by = "day")

# creation of dataframe with the dates of the previous sequence
vars_df <- data.frame(date = date_sequence)

vars_df$precipitation <- adj_pr$precipitation
vars_df$temperature <- adj_tas$temperature

```

### Resolution Adjustments

```{r}

# resolution adjustments evapotranspiration -----------------------------------

# dataframe with dates column
daily_evapotranspiration <- data.frame(date = date_sequence)

daily_evapotranspiration <- merge(daily_evapotranspiration, evspsbl, by = "date", all.x = TRUE)


daily_evapotranspiration <- daily_evapotranspiration|>
  mutate(month = format(date, "%Y-%m"))


# monthly values
monthly_values <- daily_evapotranspiration |>
  group_by(month = format(date, "%Y-%m")) |>
  summarize(monthly_evapotranspiration = mean(evapotranspiration, na.rm = TRUE))

# Fill NA values with monthly averages
filled_evapotranspiration_df <- daily_evapotranspiration |>
  left_join(monthly_values, by = c("month")) |>
  mutate(evapotranspiration = ifelse(is.na(evapotranspiration), 
                                     monthly_evapotranspiration, 
                                     evapotranspiration)) |>
  select(-month, -monthly_evapotranspiration)


# resolution adjustments radiation---------------------------------------------

# dataframe with dates column
daily_net_radiation <- data.frame(date = date_sequence)

daily_net_radiation <- merge(daily_net_radiation, net_radiation_df, by = "date", all.x = TRUE)

daily_net_radiation <- daily_net_radiation|>
  mutate(month = format(date, "%Y-%m"))


# monthly values
monthly_values <- daily_net_radiation |>
  group_by(month = format(date, "%Y-%m")) |>
  summarize(monthly_net_radiation = mean(net_radiation, na.rm = TRUE))

# Fill NA values with monthly averages
filled_net_radiation_df <- daily_net_radiation |>
  left_join(monthly_values, by = c("month")) |>
  mutate(net_radiation = ifelse(is.na(net_radiation), 
                                     monthly_net_radiation, 
                                     net_radiation)) |>
  select(-month, -monthly_net_radiation)


# add adjusted variables to the dataframe

vars_df$evapotranspiration <- filled_evapotranspiration_df$evapotranspiration
vars_df$rnet <- filled_net_radiation_df$net_radiation

# check for na values in the dataframe
colSums(is.na(vars_df))


```

### PET Calculation convert_et

evapotranspiration is already obtained in mass units. With the following function also the equilibrium evapotranspiration (eet) gets converted into mass units. The result is calculated \* 1.26 to obtain the potential evapotranspiration.

```{r}

# höheninfo modell grid

# calculate patm: https://github.com/geco-bern/rpmodel/blob/master/R/subroutines.R

elv <- 0

patm <- 101325

#patm <- calc_patm(0)

# calculate equilibrium evapotranspiration (eet) in mass units
vars_df$eet <- apply(vars_df[, c("temperature", "rnet")], 1, function(x) convert_et(x[1], x[2], patm)) # returns the potential water mass flux in mm d-1 (eqv. to kg m-2)

# remove energy units row from dataframe
vars_df <- select(vars_df, -rnet)

# calculate factor 1.26 * eet
vars_df$potential_evapotranspiration <- vars_df$eet * 1.26

```


### PET Calculation bigleaf

```{r}

## calculate equilibrium evapotranspiration (eet) using the equilibrium.ET function
  vars_df$eet <- bigleaf::equilibrium.ET(
    Tair = vars_df$temperature,
    Rn = vars_df$rnet
  )

  ## convert from energy units (W/m²) to mass units (mm/day)
  ### 1 W/m² is approximately equal to 0.0354 mm/day (latent heat flux conversion)
  eet_mm_day <- eet * 0.0354

  ## remove energy units row from dataframe
  vars_df <- select(vars_df, -rnet)

  ## calculate factor 1.26 * eet
  vars_df$potential_evapotranspiration <- vars_df$eet * 1.26

```

## Annual totals of evapotranspiration, potential evapotranspiration and precipitation

```{r, warning=FALSE}

ann_et_pet_pr <- vars_df |>
  mutate(year = year(date)) |>
  group_by(year) |>
  summarise(et = sum(evapotranspiration), pet = sum(potential_evapotranspiration), prec = sum(precipitation))

# remove year 2014 because data not available for whole year
ann_et_pet_pr <- ann_et_pet_pr[-1, ]

ann_et_pet_pr |>
  pivot_longer(cols = c(et, pet, prec), names_to = "Flux") |> 
  ggplot(aes(x = year, y = value, color = Flux)) +
  geom_line() +
  ggtitle("Annual totals of et, pet and prec")+
  labs(y = "Flux (mm/yr)")

```

## Snow Simulation

Simulation of snow accumulation and melt based on temperature and precipitation.

```{r}

vars_df <- vars_df |>
  mutate(precipitation = ifelse(temperature < 0, 0, precipitation),
         snow = ifelse(temperature < 0, precipitation, 0)) |>
  simulate_snow(varnam_prec = "precipitation", varnam_snow = "snow", varnam_temp = "temperature")


vars_df <- vars_df |>
  mutate(wbal = liquid_to_soil - evapotranspiration, wbal_pet = liquid_to_soil - potential_evapotranspiration)


plot_et <- vars_df |>
  ggplot(aes(date, evapotranspiration)) +
  geom_line()

plot_liquid_to_soil <- vars_df |>
  ggplot(aes(date, liquid_to_soil)) +
  geom_line()

plot_wbal <- vars_df |>
  ggplot(aes(date, wbal)) +
  geom_line()

plot_et
plot_liquid_to_soil
plot_wbal


```

## Cumulative Deficit Algorithm

```{r}

# calculate cumulative water deficit
out_cwd <- cwd(vars_df,
               varname_wbal = "wbal",
               varname_date = "date",
               thresh_terminate = 0.0,
               thresh_drop = 0.0)

# calculate potential cumulative water deficit
out_pcwd <- cwd(vars_df,
               varname_wbal = "wbal_pet",
               varname_date = "date",
               thresh_terminate = 0.0,
               thresh_drop = 0.0)

```

Retain only events of a minimum length of 20 days.

```{r}

out_cwd$inst <- out_cwd$inst |>
  filter(len >= 20)

out_pcwd$inst <- out_pcwd$inst |>
  filter(len >= 20)

```

```{r}

ggplot() +
  geom_rect(
    data = out_cwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_cwd$df, aes(date, precipitation), size  =  0.3, color = "royalblue") +
  geom_line(data  =  out_cwd$df, aes(date, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 200)) +
  theme_classic() +
  labs(x = "Date", y = "Cumulative water deficit (mm)")


```

this plot means that from 2020 on Niger will face a continuing water deficit. Potential cumulative water deficit should be higher than cwd.

```{r}

ggplot() +
  geom_rect(
    data = out_pcwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_pcwd$df, aes(date, precipitation), size  =  0.3, color = "royalblue") +
  geom_line(data  =  out_pcwd$df, aes(date, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 200)) +
  theme_classic() +
  labs(x = "Date", y = "Potential cumulative water deficit (mm)")

```

# Literature

Air Pressure \| National Oceanic and Atmospheric Administration: <https://www.noaa.gov/jetstream/atmosphere/air-pressure>, last access: 23 March 2024.

Brönnimann, S. P. D.: Klimatologie, 1st ed., utb GmbH, Stuttgart, 320 pp., <https://doi.org/10.36198/9783838548197>, 2018.
