---
title: "Gridpoint Extraction"
author: "Patricia Gribi"
date: "2024-04-29"
output: html_document
---


## Overview

explain what happens here...


## Libraries Loading

```{r, echo=FALSE}

# load libraries
source("../R/load_libraries.R")
load_libraries()

```


## Data Reading

**Reproducible workflow**: The file path should be adjusted to indicate the location where the data is downloaded within the local repository. Instructions for data download can be found in the 'data-raw' folder of this repository. Alternatively, the data can be obtained by cloning the following repository on GitHub: https://github.com/geco-bern/CMIP6ng_download, following the provided instructions.

```{r, echo=FALSE}

# to read the data from a local repository, adjust the file path following the nc_open("your/local/data/download/data.nc") function call to specify the location where the data is stored after download

# evapotranspiration (kg m-2 s-1)
evspsbl_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/evspsbl/mon/native/evspsbl_mon_CESM2_ssp585_r1i1p1f1_native.nc") 

# precipitation (kg m-2 s-1)
pr_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/pr/day/native/pr_day_CESM2_ssp585_r1i1p1f1_native.nc")

# air temperature (K)
tas_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/tas/day/native/tas_day_CESM2_ssp585_r1i1p1f1_native.nc")

# surface upwelling longwave radiation (W m-2)
rlus_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rlus/mon/native/rlus_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# surface downwelling longwave radiation (W m-2)
rlds_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rlds/mon/native/rlds_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# surface downwelling shortwave radiation (W m-2)
rsds_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rsds/mon/native/rsds_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# surface upwelling shortwave radiation (W m-2)
rsus_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rsus/mon/native/rsus_mon_CESM2_ssp585_r1i1p1f1_native.nc")

```


## Grid Information

To gain insights into the grid structure delineated in the CMIP6 data, an examination was conducted on the netCDF file with the precipitation variable as a representative case. By printing the dataset useful information was obtained. Within the dimensions section, it became clear that the latitude is represented as double, measured in degrees north, spanning from -90 to 90. Correspondingly, the longitude is also expressed as double, denominated in degrees east, ranging from 0 to 360. The longitudinal values are aligned along the x-axis, while the latitudinal values are aligned along the y-axis. 


```{r, echo=FALSE}

# get information about the grid
print(pr_CMIP6ng_data)
#print(evspsbl_CMIP6ng_data)

# List all variables in the NetCDF file
print(pr_CMIP6ng_data$var)

# List all dimensions in the NetCDF file
print(pr_CMIP6ng_data$dim)

nc_stack <- stack("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/evspsbl/mon/native/evspsbl_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# Print the stack to see its structure
print(nc_stack)

```


## Grid Conversion

The function preserves the position of the prime meridian (0 degrees longitude) and ensures that longitudes within the range of 0 to 180 degrees remain unchanged as well. Any negative longitudes are shifted to the equivalent positive values in the 0 to 360-degree range. 

```{r}

# function to convert longitudes from -180-180 range to 0-360 range 

convert_longitude_reverse <- function(lon) {
  lon <- lon + 360 * (lon < 0)
  return(lon)
}

# example values
lon <- c(0, -1, -179, -62)
converted_lon <- convert_longitude_reverse(lon)
print(converted_lon)

```


## Gridpoint Extraction 

The gridpoint_extraction function was implemented to retrieve a specific grid point from the supplied dataset. Consequently, it facilitates the expression of the variable at a designated spatial location across time. The functionality of this function encompasses the provision of longitude and latitude as parameters. When these parameters are not provided to the function, default values are employed. In this function, the default parameter corresponds to the coordinates of the GIUB, specifically 46.95286466584806 latitude and 7.4351598698922405 longitude.


```{r, echo=FALSE}

# check dimension order
# print(evspsbl_CMIP6ng_data$dim)

source("../R/gridpoint_extraction.R")

# evapotranspiration extraction
evspsbl <- gridpoint_extraction(evspsbl_CMIP6ng_data, "evspsbl")

# precipitation extraction
pr <- gridpoint_extraction(pr_CMIP6ng_data, "pr")

# temperature extraction
tas <- gridpoint_extraction(tas_CMIP6ng_data, "tas")

# radiation gridpoint extraction
rlus <- gridpoint_extraction(rlus_CMIP6ng_data, "rlus")
rlds <- gridpoint_extraction(rlds_CMIP6ng_data, "rlds")
rsds <- gridpoint_extraction(rsds_CMIP6ng_data, "rsds")
rsus <- gridpoint_extraction(rsus_CMIP6ng_data, "rsus")

```


## CSV Creation

```{r}

## save extracted data to files and add them to git repo

# number of rows included in the CSV file
# num_rows <- 100  
# selection of the specified number of rows
# evspsbl_subset <- head(evspsbl, num_rows)

# write data to CSV
write.csv(evspsbl, "../data/gridpoint_evspsbl.csv", row.names = FALSE)
write.csv(pr, "../data/gridpoint_pr.csv", row.names = FALSE)
write.csv(tas, "../data/gridpoint_tas.csv", row.names = FALSE)
write.csv(rlus, "../data/gridpoint_rlus.csv", row.names = FALSE)
write.csv(rlds, "../data/gridpoint_rlds.csv", row.names = FALSE)
write.csv(rsds, "../data/gridpoint_rsds.csv", row.names = FALSE)
write.csv(rsus, "../data/gridpoint_rsus.csv", row.names = FALSE)

```
