---
title: "Proposal: Implementation"
author: "Patricia Gribi"
date: "2024-03-06"
output: html_document
---

What approach will you take to achieve the goal in terms of data and methods?


# The CMIP6-ng Data

## Data Structure


```{r}

library(terra)

r <- rast("/data/scratch/CMIP6ng/cmip6-ng/pr/day/native/pr_day_CESM2_historical_r1i1p1f1_native.nc") #, lyrs = "pr_1")
print(r)

# time(r) from when to when
# names(r) names of the layers


```
3 dimensional, rows and cols representing longitude and latitude so a point on the grid,layers represent time, time ranges from 01.01.1850 until 31.12.2014. 

what is meant by resolution?

extent: special grid because only positive longitude
therefore transformations

Raster data: The basic raster model represents geographic (2D) continuous data as a two-dimensional array, where each position has a geographic (x, y) coordinate, a cell size (or resolution) and a given extent. In most geographic applications, coordinates are referenced and correspond to a geographic position, e.g., a particular latitude and longitude. Often, the model is expanded with a time dimension, stacking various two-dimensional arrays into a three-dimensional array.


look at cwd algorithm and if units are changed into desired ones.


## First Visualizations


```{r}

plot(r)

```



## Description and Variable Selection

A large part of this work consists of calculating globally the cumulative water 
deficits using the cwd-algorithm already developed by the GECO-group. The time 
series obtained provide the basis for the following trend analyses.
Certain variables are required for this purpose. These are provided by the 
CMIP6-ng data set (Brunner et al., 2020).

In order to create representative time series, the data should be available at 
daily resolution. The desired units are mm day-1.


The following variables have already been downloaded on the workstation 2: 

- **pr**: **Precipitation**, is available per day, on the native grid with units 
kg m-2 s-1
- **rlds**: **Surface Downwelling Longwave Radiation**, monthly available, on 
the native grid with units W m-2
- **tas**: **Near-Surface Air Temperature**, daily available, as well on the 
native grid, in K
- **tran**: **Transpiration**, is available per month, on the native grid with 
units kg m-2 s-1

Native grid: either data reported on a model's native grid, regridded data
reported on the data provider's preferred target grid, regridded data reported 
on a grid other than the native grid and other than the preferred target grid
or global mean data. Priorities descending. (p.4)


Variables desired:

- **Evapotranspiration (ET)** or, if ET is not available, **latent heat flux**
- **Precipitation** (for rain and snowfall)
- **Potential Evapotranspiration (PET)**. If not available, **daily surface net
radiation and temperature** (daily mean)


Variables missing:
- Evapotranspiration or latent heat flux
- Potential Evapotranspiration or daily surface net radiation


Variable Selection Suggestions: 
- Evapotranspiration: take sum of Transpiration and Evaporation
- Potential Evapotranspiration or daily surface net radiation??



# Methods

What Methods will I use?
für Trendanalyse was nötig?




