---
title: "Proposal: Implementation"
author: "Patricia Gribi"
date: "2024-03-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# The CMIP6-ng Data

## Data Structure

```{r}

library(terra)

r <- rast("/data/scratch/CMIP6ng/cmip6-ng/pr/day/native/pr_day_CESM2_historical_r1i1p1f1_native.nc") #, lyrs = "pr_1")
print(r)

# time(r) from when to when
# names(r) names of the layers


```

To give an example about the structure of the cmip6 data the variable
precipitation was loaded in r. The cmip6-ng data is raster data, whose
data type is a 3-dimensional array and in this example consisting of 192
rows, 288 columns, and 60225 layers. The rows and columns correspond to
longitude and latitude, respectively, while the layers denote distinct
temporal intervals. In this data set each layer spans one day. The
temporal sequence is delineated along the time-row including the period
from January 1, 1850, to December 31, 2014.

The row resolution indicates the spatial resolution of the raster data.
The grid cells measure 1.25 units in the x-direction and 0.9424084 units
in the y-direction.

The raster's extent defines the spatial boundaries in terms of minimum
and maximum values for longitude (x) and latitude (y). This data set
covers a global extent. A grid transformation will have to be conducted
eventually, as the current format exclusively accommodates positive
values. In non-climatic models, it is conventional to define longitude
within the range of -180 to 180 and latitude within -90 to 90.

The unit of measurement for the values is kilograms per square meter per
second.

## Visualizations

```{r}

plot(r)

```
