---
title: "One Gridpoint"
author: "Patricia Gribi"
date: "2024-03-19"
output: html_document
---


# Cumulative Water Deficits over Time for one Gridpoint


## Data Wrangling


### Overview

The data wrangling started with reading the data and loading the required libraries. The CMIP6ng data structure is organised in such a way that each variable is contained in a separate file. Therefore, each variable had to be read in individually. The data format is netcdf and contains a global grid. However, this is kept in a purely positive format (often preferred for climate data). A grid point was extracted for every variable. Also some unit conversions had to be performed. Variables such as precipitation and evapotranspiration were converted to the desired mm day-1 format (equivalent to kg m-2). The temporal resolution of the individual variables is daily or monthly. Linear interpolation was therefore applied to the variables that are only available monthly in order to obtain the same resolution for all variables. Then, evapotranspiration and potential evapotranspiration were calculated over time for a specific grid point. Furthermore, snow was modelled. Thanks to all these steps, the cwd algorithm could finally be applied to the data.


### Library Loading

```{r, echo=FALSE}

# libraries

library(terra)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
library(viridisLite)
library(weathermetrics)
library(ncdf4)
library(chron)
library(RColorBrewer)
library(lattice)
library(cwd)
library(lubridate)
library(rpmodel)

```


### Data Reading

```{r, echo=FALSE}

# evapotranspiration (kg m-2 s-1)
evspsbl_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/evspsbl/mon/native/evspsbl_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# precipitation (kg m-2 s-1)
pr_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/pr/day/native/pr_day_CESM2_ssp585_r1i1p1f1_native.nc") 

# air temperature (K)
tas_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/tas/day/native/tas_day_CESM2_ssp585_r1i1p1f1_native.nc")

# surface upwelling longwave radiation (W m-2)
rlus_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rlus/mon/native/rlus_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# surface downwelling longwave radiation (W m-2)
rlds_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rlds/mon/native/rlds_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# surface downwelling shortwave radiation (W m-2)
rsds_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rsds/mon/native/rsds_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# surface upwelling shortwave radiation (W m-2)
rsus_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rsus/mon/native/rsus_mon_CESM2_ssp585_r1i1p1f1_native.nc")

```


### Grid Information

To gain insights into the grid structure delineated in the CMIP6 data, an examination was conducted on the netCDF file with the precipitation variable as a representative case. By printing the dataset useful information was obtained. Within the dimensions section, it became clear that the latitude is represented as double, measured in degrees north, spanning from -90 to 90. Correspondingly, the longitude is also expressed as double, denominated in degrees east, ranging from 0 to 360. The longitudinal values are aligned along the x-axis, while the latitudinal values are aligned along the y-axis. Consequently, in coordinate notation (x, y), the longitude assumes the primary position, followed by the latitude.


```{r}

# get information about the grid
print(pr_CMIP6ng_data)

```


### Gridpoint Extraction 

The gridpoint_extraction function was implemented to retrieve a specific grid point from the supplied dataset. Consequently, it facilitates the expression of the variable at a designated spatial location across time. The functionality of this function encompasses the provision of longitude and latitude as parameters. When these parameters are not provided to the function, default values are employed. In this function, the default parameter corresponds to the coordinates of the GIUB, specifically 46.95286466584806 latitude and 7.4351598698922405 longitude.

```{r, echo=FALSE}

source("../R/gridpoint_extraction.R")

# evapotranspiration extraction
evspsbl <- gridpoint_extraction(evspsbl_CMIP6ng_data, "evspsbl")

# precipitation extraction
pr <- gridpoint_extraction(pr_CMIP6ng_data, "pr")

# temperature extraction
tas <- gridpoint_extraction(tas_CMIP6ng_data, "tas")

# radiation gridpoint extraction
rlus <- gridpoint_extraction(rlus_CMIP6ng_data, "rlus")
rlds <- gridpoint_extraction(rlds_CMIP6ng_data, "rlds")
rsds <- gridpoint_extraction(rsds_CMIP6ng_data, "rsds")
rsus <- gridpoint_extraction(rsus_CMIP6ng_data, "rsus")

```


### Grid Conversion

This function serves the purpose of determining the position of a point within the standardized grid. Its utility extends to the conversion of longitude values exceeding 180 degrees into their corresponding negative equivalents.

```{r}

# Function to convert longitudes from 0-360 range to -180-180 range
convert_longitude <- function(lon) {
  lon <- lon - 360 * (lon > 180)
  return(lon)
}

# Default value of gridpoint_extraction function:
lon <- 340
converted_lon <- convert_longitude(lon)
print(converted_lon)

```

### Column Renaming and Unit Conversions

This research project aims to consistently represent the variables
across all data sets with the following units:

- **Date**: year-month-day (xxxx-xx-xx)
- **Precipitation** and **(potential) Evapotranspiration**: in mm day-1 (equivalent to kg m-2)
- **Temperature (tas)**: in degree Celsius (°C)
- **Atmospheric Pressure**: in Pascal (Pa)
- **Radiation**: in W m-2

Therefore some unit conversions had to be conducted.

```{r}

# evapotranspiration
colnames(evspsbl) <- c("date", "evapotranspiration") # column renaming
evspsbl$evapotranspiration <- evspsbl$evapotranspiration * 86400 # conversion to mm day-1

# precipitation
colnames(pr) <- c("date", "precipitation") # column renaming
pr$precipitation <- pr$precipitation * 86400 # conversion to mm day-1

# temperature
colnames(tas) <- c("date", "temperature") # column renaming
tas$temperature <- tas$temperature - 273.15 # conversion to °C

# radiation
colnames(rlus) <- c("date", "up_longwave_radiation") # column renaming
colnames(rlds) <- c("date", "down_longwave_radiation") # column renaming
colnames(rsds) <- c("date", "down_shortwave_radiation") # column renaming
colnames(rsus) <- c("date", "up_shortwave_radiation") # column renaming

# merge radiation variables into one dataframe
radiation_df <- merge(merge(merge(rlus, rlds, by = "date"), rsds, by = "date"), rsus, by = "date")

# calculate net radiation and add to dataframe
radiation_df$net_radiation <- (radiation_df$down_shortwave_radiation - radiation_df$up_shortwave_radiation) + (radiation_df$down_longwave_radiation - radiation_df$up_longwave_radiation)

```

In order to calculate the potential evapotranspiration the net radiation was needed. The net radiation is calculated as net shortwave radiation + net longwave radiation (Citation). The approach in this thesis considers the potential evapotranspiration as 1.26 * the net radiation (Citation). 


### Time-Range and Resolution Adjustments

To obtain a daily resolution for all variables, linear interpolation was applied to monthly available variables.

```{r}

# creation of new dataframe to calculate pet

source("../R/exclude_time_range.R")

# precipitation subset selection
adj_pr <- exclude_time_range(pr, as.Date("2014-11-22"), as.Date("2014-12-07") )
adj_pr <- exclude_time_range(adj_pr,as.Date("2100-10-17"), as.Date("2100-11-01") )

# temperature subset selection
adj_tas <- exclude_time_range(tas, as.Date("2014-11-22"), as.Date("2014-12-07") )
adj_tas <- exclude_time_range(adj_tas,as.Date("2100-10-17"), as.Date("2100-11-01") )

# creation of the date sequence
date_sequence <- seq(as.Date("2014-12-07"), as.Date("2100-10-16"), by = "day")

# creation of dataframe with the dates of the previous sequence
vars_df <- data.frame(date = date_sequence)

vars_df$precipitation <- adj_pr$precipitation
vars_df$temperature <- adj_tas$temperature

# addition of evapotranspiration with linear interpolation to obtain daily resolution
vars_df$evapotranspiration <- approx(x= evspsbl$date, y = evspsbl$evapotranspiration, xout = vars_df$date)$y

# addition of net radiation with linear interpolation to obtain daily resolution
vars_df$eet_energy_units <- approx(x= radiation_df$date, y = radiation_df$net_radiation, xout = vars_df$date)$y

```


### PET

evapotranspiration is already in mass units. With the following function also the potential evapotranspiration gets converted into mass units.

```{r}

# höheninfo modell grid

# calculate patm: https://github.com/geco-bern/rpmodel/blob/master/R/subroutines.R

elv <- 0

patm <- 101325

#patm <- calc_patm(0)

# calculate potential evapotranspiration (pet)
vars_df$eet <- apply(vars_df[, c("temperature", "eet_energy_units")], 1, function(x) convert_et(x[1], x[2], patm)) # returns the potential water mass flux in mm d-1 (eqv. to kg m-2)

# remove energy units row from dataframe
vars_df <- select(vars_df, -eet_energy_units)

# calculate factor 1.26 * eet
vars_df$potential_evapotranspiration <- vars_df$eet * 1.26

```


## Analysis

### Comparison of the evapotranspiration and the potential evapotranspiration

```{r}

ann_evspsbl_et <- vars_df |>
  mutate(year = year(date)) |>
  group_by(year) |>
  summarise(et = sum(evapotranspiration), pet = sum(potential_evapotranspiration))

ann_evspsbl_et |>
  pivot_longer(cols = c(et, pet), names_to = "Flux") |> 
  ggplot(aes(x = year, y = value, color = Flux)) +
  geom_line() +
  labs(y = "Flux (mm/yr)")

```
can the actual evapotranspiration be higher than the potential evapotranspiration? Why are there parts where this is the case?



### Annual totals of evapotranspiration, potential evapotranspiration and precipitation

The evspsbl variable was chosen, as it is a direct variable of the CMIP6ng data and was not calculated indirectly as with the evapotranspiration variable. Which was obtained from the information on pressure, latent heat flux and temperature. As observable in the plot below the data matches very well with the precipiation.

```{r, warning=FALSE}

ann_et_pet_pr <- vars_df |>
  mutate(year = year(date)) |>
  group_by(year) |>
  summarise(et = sum(evapotranspiration), pet = sum(potential_evapotranspiration), prec = sum(precipitation))

# remove year 2014 because data not available for whole year
ann_et_pet_pr <- ann_et_pet_pr[-1, ]

ann_et_pet_pr |>
  pivot_longer(cols = c(et, pet, prec), names_to = "Flux") |> 
  ggplot(aes(x = year, y = value, color = Flux)) +
  geom_line() +
  labs(y = "Flux (mm/yr)")

```
explain this figure

add other values and plot again

20 long, 30 lat, curves et and prec were almost the same
water-limited system, runoff probably small or constant, no long-term change in storage


## Snow Simulation

```{r}

vars_df <- vars_df |>
  mutate(precipitation = ifelse(temperature < 0, 0, precipitation),
         snow = ifelse(temperature < 0, precipitation, 0)) |>
  simulate_snow(varnam_prec = "precipitation", varnam_snow = "snow", varnam_temp = "temperature")


vars_df <- vars_df |>
  mutate(wbal = liquid_to_soil - evapotranspiration, wbal_pet = liquid_to_soil - potential_evapotranspiration)

```


## Cumulative Deficit Algorithm

```{r}

# calculate cumulative water deficit
out_cwd <- cwd(vars_df,
               varname_wbal = "wbal",
               varname_date = "date",
               thresh_terminate = 0.0,
               thresh_drop = 0.0)

# calculate potential cumulative water deficit
out_pcwd <- cwd(vars_df,
               varname_wbal = "wbal_pet",
               varname_date = "date",
               thresh_terminate = 0.0,
               thresh_drop = 0.0)

```

Retain only events of a minimum length of 20 days.

```{r}

out_cwd$inst <- out_cwd$inst |>
  filter(len >= 20)

out_pcwd$inst <- out_pcwd$inst |>
  filter(len >= 20)

```



```{r}

ggplot() +
  geom_rect(
    data = out_cwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_cwd$df, aes(date, precipitation), size  =  0.3, color = "royalblue") +
  geom_line(data  =  out_cwd$df, aes(date, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 200)) +
  theme_classic() +
  labs(x = "Date", y = "Cumulative water deficit (mm)")


```
this result is strange as in the plot above it was shown that the precipitation is always higher than the evapotranspiration. How can it be that the deficit is higher now than the precipitation?


```{r}

ggplot() +
  geom_rect(
    data = out_pcwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_pcwd$df, aes(date, precipitation), size  =  0.3, color = "royalblue") +
  geom_line(data  =  out_pcwd$df, aes(date, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 200)) +
  theme_classic() +
  labs(x = "Date", y = "Potential cumulative water deficit (mm)")

```




# Literature

Air Pressure | National Oceanic and Atmospheric Administration: https://www.noaa.gov/jetstream/atmosphere/air-pressure, last access: 23 March 2024.






