---
title: "One Gridpoint"
author: "Patricia Gribi"
date: "2024-03-19"
output: html_document
---


9.10-10.40
11.00-

# Variables over Time for one Gridpoint


```{r}

# libraries

library(terra)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
library(viridisLite)
library(weathermetrics)
library(ncdf4)
library(chron)
library(RColorBrewer)
library(lattice)
library(cwd)

```


# Evapotranspiration

## Variables


```{r}

# latent heat flux (W m-2)
hfls_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/hfls/mon/native/hfls_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# air temperature (°C)
tas_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/tas/day/native/tas_day_CESM2_ssp585_r1i1p1f1_native.nc") 

# atmospheric pressure (Pa)

patm <- 101320 # standard air pressure at sea level in Pa (1013.2 mb) (Air Pressure | National Oceanic and Atmospheric Administration, 2024)

```


## Extraction

```{r}

gridpoint_extraction <- function(data, variable) {
  
  # indices of the grid point 
  start_lon <- 20  # longitude index
  start_lat <- 30 # latitude index
  
  # get days
  var_dates <- as.Date(data$dim$time$vals, origin = '1850-01-01')
  
  # get values at location lon lat
  var_output <- ncvar_get(data, varid = variable,
                    start= c(which.min(abs(data$dim$lon$vals - lon)), # look for closest long
                             which.min(abs(data$dim$lat$vals - lat)),  # look for closest lat
                             1),
                    count = c(1,1,-1)) #count '-1' means 'all values along that dimension'that dimension'
  
  # create dataframe
  var <- data.frame(dates= var_dates, variable = var_output)
  
  return(var)
}


tas <- gridpont_extraction(tas_CMIP6ng_data, "tas")



```


## Conversion into mass units

```{r}

et <- convert_et(hfls_CMIP6ng_data, tas_CMIP6ng_data, patm) # returns the water mass flux in mm d-1 (eqv. to kg m-2)


```

I have different variables in netcdf format
i want to extract only the variables of the netcdf data, combine them to obtain just one variable (evapotranspiration) and then put it in the same netcdf-format (regarding longitude/latitude, extent...). How can I do this?


Create a New NetCDF File: Create a new NetCDF file using the nc_create() function. Define the dimensions (longitude, latitude, time, etc.) and attributes based on the original NetCDF files.

Write Combined Variable to the New NetCDF File: Write the combined variable (evapotranspiration) to the new NetCDF file using ncvar_put().


```{r}

# Create a new NetCDF file
new_nc_file <- nc_create("combined_evapotranspiration.nc")

# Define dimensions and attributes based on the original NetCDF files
# (Add code to define dimensions and attributes)

# Write evapotranspiration to the new NetCDF file
ncvar_put(new_nc_file, "evapotranspiration", evapotranspiration)


```





## Gridpoint extraction

```{r}

# loading data
pr_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/pr/day/native/pr_day_CESM2_ssp585_r1i1p1f1_native.nc") 

#print(pr_CMIP6ng_data) # dimension-check

# indices of the grid point 
start_lon <- 20  # longitude index
start_lat <- 30 # latitude index

# get days
pr_dates <- as.Date(pr_CMIP6ng_data$dim$time$vals, origin = '1850-01-01')

# get values at location lon lat
pr_output <- ncvar_get(pr_CMIP6ng_data, varid = 'pr',
                  start= c(which.min(abs(pr_CMIP6ng_data$dim$lon$vals - lon)), # look for closest long
                           which.min(abs(pr_CMIP6ng_data$dim$lat$vals - lat)),  # look for closest lat
                           1),
                  count = c(1,1,-1)) #count '-1' means 'all values along that dimension'that dimension'

# create dataframe
pr <- data.frame(dates= pr_dates, precipitation = pr_output)


```



```{r}

ggplot(data = pr, aes(x = dates, y = precipitation)) +
  geom_line() +
  labs(x = "Date", y = "Precipitation", title = "Precipitation over Time at Specific Grid Point") +
  theme_minimal()

```


find out what gridpoint?
wrong dates of data
convert units of precipitation
how to run the variables through the cwd algorithm in another repository


```{r}

nc_close(pr)


# plot

var <- tmp.array[,,1]

mapCDFpr <- function(lat, lon, var)
  
{
  
  titletext <- "Precipitation"
  
  expand.grid(lon, lat) %>%
    rename(longitude = lon, latitude = lat) %>%
    
    mutate(lon = ifelse(lon > 180, -(360-lon), lon),
           
           var = as.vector(var)) %>%
    
    ggplot() +
    
    geom_point(aes(x = lon, y = lat, color = var),
               size = 0.8) +
    borders("world", colour = "black", fill = NA) +
    
    scale_color_viridis(na.value = "white", name = "Precipitation") + 
    
    theme(legend.direction ="vertical", legend.position = "right", legend.key.width = unit(0.4, "cm"),    legend.key.height = unit(2, "cm")) + 
    
    coord_quickmap() +
    
    ggtitle(titletext)
  
}


mapCDFpr(lat, lon, tmp.slice)


```





```{r}

# extraction of a specific grid point 
lon <- 80
lat <- 0
pr_grid_point <- extract(pr, c(lon, lat))
print(pr_grid_point)

```

# Literature

Air Pressure | National Oceanic and Atmospheric Administration: https://www.noaa.gov/jetstream/atmosphere/air-pressure, last access: 23 March 2024.






