---
title: "One Gridpoint"
author: "Patricia Gribi"
date: "2024-03-19"
output: html_document
---


# Variables over Time for one Gridpoint


## Data and Library Loading


```{r, echo=FALSE}

# libraries

library(terra)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
library(viridisLite)
library(weathermetrics)
library(ncdf4)
library(chron)
library(RColorBrewer)
library(lattice)
library(cwd)
library(lubridate)

```



```{r}

# latent heat flux (W m-2)
hfls_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/hfls/mon/native/hfls_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# air temperature (K)
tas_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/tas/day/native/tas_day_CESM2_ssp585_r1i1p1f1_native.nc") 

# precipitation (kg m-2 s-1)
pr_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/pr/day/native/pr_day_CESM2_ssp585_r1i1p1f1_native.nc") 

# atmospheric pressure (Pa)
patm <- 101320 # standard air pressure at sea level in Pa (1013.2 mb) (Air Pressure | National Oceanic and Atmospheric Administration, 2024)

# surface upwelling longwave radiation (W m-2)
rlus_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rlus/mon/native/rlus_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# surface downwelling longwave radiation (W m-2)
rlds_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rlds/mon/native/rlds_mon_CESM2_ssp585_r1i1p1f1_native.nc")

# surface downwelling shortwave radiation (W m-2)
rsds_CMIP6ng_data <- nc_open("/home/patricia/data_download/CMIP6ng_download/cmip6-ng/rsds/mon/native/rsds_mon_CESM2_ssp585_r1i1p1f1_native.nc")

```


## Gridpoint Extraction

```{r}

source("../R/gridpoint_extraction.R")


# temperature extraction
tas <- gridpoint_extraction(tas_CMIP6ng_data, "tas")

colnames(tas) <- c("date", "temperature") # column renaming

tas$temperature <- tas$temperature - 273.15 # conversion to °C


# latent heat flux extraction

hfls <- gridpoint_extraction(hfls_CMIP6ng_data, "hfls")

colnames(hfls) <- c("date", "latent_heat_flux") # column renaming


# precipitation extraction

pr <- gridpoint_extraction(pr_CMIP6ng_data, "pr")

colnames(pr) <- c("date", "precipitation") # column renaming

```


## Time-Range and Resolution Adjustments

```{r}

# creation of new dataframe to calculate et

# temperature subset selection to match latent_heat_flux records
source("../R/exclude_time_range.R")

adj_tas <- exclude_time_range(tas, as.Date("2014-11-22"), as.Date("2014-12-07") )
adj_tas <- exclude_time_range(adj_tas,as.Date("2100-10-17"), as.Date("2100-11-01") )

# precipitation subset selection to match the other dataframes

adj_pr <- exclude_time_range(pr, as.Date("2014-11-22"), as.Date("2014-12-07") )
adj_pr <- exclude_time_range(adj_pr,as.Date("2100-10-17"), as.Date("2100-11-01") )


# creation of the date sequence
date_sequence <- seq(as.Date("2014-12-07"), as.Date("2100-10-16"), by = "day")

# creation of dataframe with the dates of the previous sequence
vars_df <- data.frame(date = date_sequence)

vars_df$temperature <- adj_tas$temperature
vars_df$precipitation <- adj_pr$precipitation

# addition of latent heat flux with linear interpolation to obtain daily resolution
vars_df$latent_heat_flux <- approx(x= hfls$date, y = hfls$latent_heat_flux, xout = vars_df$date)$y

```


## Conversion of ET into mass units

```{r}

# calculate evapotranspiration
vars_df$evapotranspiration <- apply(vars_df[, c("temperature", "latent_heat_flux")], 1, function(x) convert_et(x[1], x[2], patm)) # returns the water mass flux in mm d-1 (eqv. to kg m-2)

```


### Annual totals

```{r}

ann_et_pr <- vars_df |>
  mutate(year = year(date)) |>
  group_by(year) |>
  summarise(et = sum(evapotranspiration), prec = sum(precipitation))

# remove year 2014 because data not available for whole year
ann_et_pr <- ann_et_pr[-1, ]

ann_et_pr |>
  pivot_longer(cols = c(et, prec), names_to = "Flux") |> 
  ggplot(aes(x = year, y = value, color = Flux)) +
  geom_line() +
  labs(y = "Flux (mm/yr)")

```


## Snow Simulation

```{r}

vars_df <- vars_df |>
  mutate(precipitation = ifelse(temperature < 0, 0, precipitation),
         snow = ifelse(temperature < 0, precipitation, 0)) |>
  simulate_snow(varnam_prec = "precipitation", varnam_snow = "snow", varnam_temp = "temperature")


vars_df <- vars_df |>
  mutate(wbal = liquid_to_soil - evapotranspiration)

```


## Cumulative Deficit Algorithm

```{r}

out_cwd <- cwd(vars_df,
               varname_wbal = "wbal",
               varname_date = "date",
               thresh_terminate = 0.0,
               thresh_drop = 0.0)


```

Retain only events of a minimum length of 20 days.

```{r}

out_cwd$inst <- out_cwd$inst |>
  filter(len >= 20)

```



```{r}

ggplot() +
  geom_rect(
    data = out_cwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = -99, ymax = 99999),
    fill = rgb(0,0,0,0.3),
    color = NA) +
  geom_line(data  =  out_cwd$df, aes(date, precipitation), size  =  0.3, color = "royalblue") +
  geom_line(data  =  out_cwd$df, aes(date, deficit), color = "tomato") +
  coord_cartesian(ylim = c(0, 200)) +
  theme_classic() +
  labs(x = "Date", y = "Cumulative water deficit (mm)")


```







```{r}

ggplot(data = pr, aes(x = dates, y = precipitation)) +
  geom_line() +
  labs(x = "Date", y = "Precipitation", title = "Precipitation over Time at Specific Grid Point") +
  theme_minimal()

```



## Daily Surface Net Radiation


 surface down- and upwelling long- and shortwave radiation, as well as temperature.






# Literature

Air Pressure | National Oceanic and Atmospheric Administration: https://www.noaa.gov/jetstream/atmosphere/air-pressure, last access: 23 March 2024.






